On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-25 tex/draft/draft4.tex
section 4 edits

We now present some representative examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons made to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.  Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and the regimes of validity of our two-layer cooling model are examined in \S\ref{sec:endoftime}.

\begin{figure}[tb]
\centering
%AY: figure should probably use R_B for consistency with text.
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.pdf}
%\vspace{-0.5in}
\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (including the core) for a $5 M_{\oplus}$ core at $a=60$ AU.   Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $9.24 M_{\oplus}$, respectively.  (See text for a description of evolution to yet higher masses.)   Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.  The radial profiles extend from the core to the Hill radius.} 
\label{fig:profiles}
\end{figure}

\subsection{Atmospheric Stucture}
\label{sec:profiles}
Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth.  In these examples, we fix the core mass at $M_{\rm{c}}=5 M_{\oplus}$ and the radial disk location to $a=60$ AU.  The quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.  The 9.24 $M_{\oplus}$ solution is the highest mass we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.

@@

The detailed structure of these profiles is well approximated by our non-self-gravitating, analytic solutions.  Deep in convective zones, \Eq{eq:deep} gives that $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.  This behavior is seen in the low mass solutions in \Fig{fig:profiles}.  As explained by \Eq{eq:TPsg}, the high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$.  In agreement with \Eq{eq:cb2}, the radiative zones remain nearly isothermal, even for the higher masses.  Consequently, the pressure increases nearly exponentially with inverse depth.

@@

The cooling model of \S\ref{sec:twolayer} is used to connect solutions of different atmospheric masses into an evolutionary sequence.  \Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.  During the early stages of atmospheric growth, the luminosity drops sharply.  This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).  The increasing pressure depth of the RCB with mass causes increases in the optical depth ($\propto \kappa P)$ and decreases in the radiative luminosity.  This behavior is described in \Eqs{eq:crit}{eq:Lcb}.

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.  To understand this difference, we note that the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, holds in both cases; thus accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.   To allow these relatively lower $P\cb$ values (and thus higher luminosities) in the higher mass solutions, the radius $R\cb$ must shift outwards, as shown in \Fig{fig:profiles}.

@@

\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in our reference disk model.  This instantaneous growth time shows clearly that the atmosphere spends the bulk of its time growing though intermediate atmospheric masses, $\sim 1 -3$ $M_\oplus$ in this case.  Growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.  The accelerated growth at higher masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.

The fact that growth times peak when $M_{\rm{atm}} < M_{\rm c}$, before the crossover mass, is helpful for understanding atmospheric growth. In contrast with our analytic model, where the time to runaway accretion cannot be calculated self-consistently, the numerical model provides information about when unstable collapse occurs, i.e. when growth times decrease. It follows that the time to runaway accretion, $t_{\rm run}$, can be estimated as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to a fraction $f$ of its maximum value. This allows for a more rigorous estimate of the runaway accretion timescale, and thus of the critical core mass, as opposed to choosing $M_{\rm atm} \sim M\co$ as the criterion for collapse. In practice, however, the continuous accelerated growth in the late stages of the atmosphere evolution renders the estimated timescale to envelope collapse, and therefore the critical core mass, insensitive to the exact choice of runaway accretion time or atmosphere mass.  Since $M_{\rm atm} < M\co$ when growth times peak, we may proceed by analogy with the analytic model and set $t=t_{\rm co}$, when $M_{\rm atm}=M\co$, as the criterion for collapse. We note that in some cases our cooling model breaks down before the crossover mass, as explained in section \S\ref{sec:endoftime}, hence evolution is stopped before $M_{\rm atm}=M\co$. However, this breakdown occurs for $M_{\rm atm} \sim 0.8-0.9 M\co$, comparable to the crossover mass, so the additional time to reach crossover is negligible. It follows that $t_{\rm run} \approx t_{\rm co}$. By analogy with the analytic results we thus choose $t_{\rm co}$ as the timescale to runaway accretion, as it will give the same numerical results as $t_{\rm run}$ and it is easier to understand in the context of other core accretion models.

%The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.  This insensitivity to the upper mass threshold is characteristic of accelerating growth.

%We next show how this behavior is fortuitous, as our model assumptions can break down during the later stages of atmospheric growth.

@@

To examine the regimes where our cooling model is valid, we compare the model luminosity to the additional neglected luminosity, $L_{\rm negl}$, that a more detailed model would have generated in the radiative zone.  To compute $L_{\rm negl}$ we compute the entropy difference between successive radiative zone solutions.  We then integrate the energy equation, $\p L / \p m = - T \p S/ \p t$, over the average depth of the radiative zone.\footnote{While useful as a diagnostic, the neglected luminosity cannot reliably correct the global cooling model because the effects of $L_{\rm negl}$ on the structure of the radiative zone are still ignored.}

 \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.  However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case, meaning that our model becomes inaccurate at late states.  Nevertheless, our cooling model gives  reasonably reliable estimates of core accretion timescales.   As shown above, the growth time is mostly spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
 %Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.  
 %AY: not sure we can back up the conservative claim because of the detailed effects on structure that haven't been calculated.

The (non-vanishing) individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.  At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.  As the mass increases, the surface terms become more significant, led by the accretion energy.  However the surface terms are everywhere smaller than $L_{\rm negl}$.  Thus wherever our model is accurate -- including the crucial early phases of growth -- surface terms are a minor correction.  
%while the surface terms appear to reduce the total luminosity (compared to $-\dot{E}$), $L_{\rm negl}$ can more than compensate, showing again that our model conservatively underestimates cooling.
%AY:again don't think conservative claim can be so easily justified.

At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as negative time steps when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.  Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after growth has started to accelerate. As the breakdown always occurs after growth times peak, it does not affect our estimate of the timescale to runaway accretion $t_{\rm run}$. In our models the breakdown occurs at lower atmosphere mass as the semimajor axis increases, since accelerated and eventually runaway growth occurs sooner at larger distances, as seen in \Fig{fig:growthtime}. The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.   Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.

@@

\item Runaway growth begins roughly at the crossover mass, when atmosphere and core masses are equal, in agreement with previous work \citep{pollack96}.  For larger core masses, runaway growth begins at larger ratios of atmosphere to core mass.

\item Self-gravity affects atmospheric evolution before crossover.  Significant self-gravitational corrections appear when the atmosphere is only $\sim$$20 \%$ as massive as the core.

>>>>

We present examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.  Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and validity of our two-layer cooling model is examined in \S\ref{sec:endoftime}.

\begin{figure}[tb]
\centering
%AY: figure should probably use R_B for consistency with text.
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.pdf}
%\vspace{-0.5in}
\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $a=60$ AU.   Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $9.24 M_{\oplus}$, respectively.  (See text for a description of evolution to yet higher masses.)   Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.  The radial profiles extend from the core to the Hill radius.} 
\label{fig:profiles}
\end{figure}

\subsection{Atmospheric Stucture}
\label{sec:profiles}
Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth around $5 M_{\oplus}$ core at $60$ AU.  Quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.  The 9.24 $M_{\oplus}$ solution is the highest mass we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.

@@

The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.  Deep in convective zones, \Eq{eq:deep} gives $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.  This behavior is seen in the low mass solutions in \Fig{fig:profiles}.  As explained by \Eq{eq:TPsg}, the high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$.  In agreement with \Eq{eq:cb2}, the radiative zones remain nearly isothermal, even for the higher masses.  Consequently, the pressure increases nearly exponentially with inverse depth.

@@

The cooling model of \S\ref{sec:twolayer} is used to connect solutions of different atmospheric masses into an evolutionary sequence.  \Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.  During the early stages of atmospheric growth, the luminosity drops sharply.  This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).  With increasing atmospheric mass, the pressure depth of the RCB increases, along with the optical depth ($\propto \kappa P\cb)$.  Consequently, the radiative luminosity decreases.  This behavior is described in \Eqs{eq:crit}{eq:Lcb}.

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.

@@

\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in our reference disk model.  This instantaneous growth time shows clearly that the atmosphere spends the bulk of its time growing though intermediate atmospheric masses, $\sim 1 -3$ $M_\oplus$ in this case.  Growth times are short both early -- when the radiative zone is transparent -- and late -- when self-gravity accelerates growth.

The fact that growth times have a well defined maximum is a characteristic of runaway growth.  Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway growth begins.  Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.  \Fig{fig:growthtime} shows that further from the star runaway growth begins at smaller $M_{\rm atm}/M\co$.  \Fig{fig:tvsM} (described in the next section) shows how the onset of runaway growth depends on core mass.

We quantify the runaway growth timescale, $t_{\rm run}$, as the time when $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value.  The choice of 10\% is arbitrary; the precise threshold chosen is relatively unimportant because growth continues to accelerate.

%The onset of runaway growth at high masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.

%peak when $M_{\rm{atm}} < M_{\rm c}$, before the crossover mass, is helpful for understanding atmospheric growth. In contrast with our analytic model, where the time to runaway accretion cannot be calculated self-consistently, the numerical model provides information about when unstable collapse occurs, i.e. when growth times decrease. It follows that the time to runaway accretion, $t_{\rm run}$, can be estimated as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to a fraction $f$ of its maximum value. This allows for a more rigorous estimate of the runaway accretion timescale, and thus of the critical core mass, as opposed to choosing $M_{\rm atm} \sim M\co$ as the criterion for collapse. In practice, however, the continuous accelerated growth in the late stages of the atmosphere evolution renders the estimated timescale to envelope collapse, and therefore the critical core mass, insensitive to the exact choice of runaway accretion time or atmosphere mass.  Since $M_{\rm atm} < M\co$ when growth times peak, we may proceed by analogy with the analytic model and set $t=t_{\rm co}$, when $M_{\rm atm}=M\co$, as the criterion for collapse. We note that in some cases our cooling model breaks down before the crossover mass, as explained in section \S\ref{sec:endoftime}, hence evolution is stopped before $M_{\rm atm}=M\co$. However, this breakdown occurs for $M_{\rm atm} \sim 0.8-0.9 M\co$, comparable to the crossover mass, so the additional time to reach crossover is negligible. It follows that $t_{\rm run} \approx t_{\rm co}$. By analogy with the analytic results we thus choose $t_{\rm co}$ as the timescale to runaway accretion, as it will give the same numerical results as $t_{\rm run}$ and it is easier to understand in the context of other core accretion models.

@@

We examine the  validity of our cooling model by compare our model luminosity to the neglected luminosity, $L_{\rm negl}$,  that a more detailed model would generate in the radiative zone.  We compute $L_{\rm negl}$ from the entropy difference between successive radiative zone solutions.  We then integrate the energy equation, $\p L / \p m = - T \p S/ \p t$, over the average depth of the radiative zone.\footnote{While useful as a diagnostic, the neglected luminosity cannot reliably correct the global cooling model because the effects of $L_{\rm negl}$ on the structure of the radiative zone are still ignored.}

 \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.  However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case.  Our cooling model is thus inaccurate at higher masses.  However the model remains reasonably accurate up to the beginning stages of runaway growth, which is sufficient for our purposes of widely exploring parameter space and exploring trends.

The individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.  At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.  As the mass increases, the surface terms become more significant, led by the accretion energy.  However the surface terms are everywhere smaller than $L_{\rm negl}$.  Thus wherever our model is accurate -- including the crucial early phases of growth -- surface terms are a minor correction.  The neglect of surface terms in the analytic model is thus not a serious omission.

At even higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity overwhelms the model luminosity by this point.    Effectively the model is telling us when to ignore it.  Our results are unaffected by the breakdown, which occurs comfortably after $t_{\rm run}$, our threshold for runaway growth.

@@

\item Runaway growth begins roughly at the crossover mass, when atmosphere and core masses are equal, $M_{\rm atm} \sim M\co$, in agreement with previous work \citep{pollack96}.  Further from the star, runaway growth begins at smaller $M_{\rm atm} \sim M\co$ ratios.  For larger core masses, runaway growth begins at larger values of $M_{\rm atm} \sim M\co$.

\item Self-gravity affects atmospheric evolution before crossover.  Significant self-gravitational corrections appear when the atmosphere is only $\sim$$10 \%$ as massive as the core.
